---
title: "Data Frame Manipulation with dplyr"
subtitle: "Adapted from Software Carpentry"
---

## Overview

Today we will learn to:

-   Use the six main data frame manipulation 'verbs' with pipes in dplyr
-   Understand how `group_by()` and `summarize()` can be combined to summarize datasets
-   Analyze a subset of data using logical filtering

## Questions

-   How can I manipulate data frames without repeating myself?

## The Problem with Base R

We often need to select observations, group data, or calculate statistics.

Using base R:

```{r}
#| eval: false
#| echo: true

gapminder <- read.csv(
  "https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/main/episodes/data/gapminder_data.csv"
)

mean(gapminder$gdpPercap[gapminder$continent == "Africa"])
mean(gapminder$gdpPercap[gapminder$continent == "Americas"])
mean(gapminder$gdpPercap[gapminder$continent == "Asia"])
```

This is repetitive and error-prone!

## The dplyr Package

The `dplyr` package provides functions for manipulating data frames that:

-   Reduce repetition
-   Reduce probability of errors
-   Are easier to read

## Tidyverse {.callout-tip}

`dplyr` belongs to the **Tidyverse** - a family of R packages designed for data science.

These packages work harmoniously together. Learn more at: [tidyverse.org](https://www.tidyverse.org/)

## The Five Main Verbs

We'll cover these commonly used functions:

1.  `select()` - choose columns
2.  `filter()` - choose rows
3.  `group_by()` - group data
4.  `summarize()` - calculate summaries
5.  `mutate()` - create new columns

Plus the pipe operator: `%>%`

## Loading dplyr

```{r}
#| eval: false
#| echo: true
#| message: false

library(dplyr)
```

If not installed, run: `install.packages("dplyr")`

## Using select()

Keep only specific columns:

```{r}
#| eval: false
#| echo: true

year_country_gdp <- select(gapminder, year, country, gdpPercap)
head(year_country_gdp)
```

## Removing Columns with select()

Use `-` to remove columns:

```{r}
#| eval: false
#| echo: true

smaller_gapminder <- select(gapminder, -continent)
head(smaller_gapminder)
```

## Introduction to Pipes

The pipe symbol `%>%` passes data from one function to the next:

```{r}
#| eval: false
#| echo: true

year_country_gdp <- gapminder %>% 
  select(year, country, gdpPercap)

head(year_country_gdp)
```

## Understanding Pipes

Step by step:

1.  Start with `gapminder` data frame
2.  Pass it using `%>%` to the next step
3.  `select()` receives the data automatically

**Fun Fact:** Similar to the shell pipe `|` but the concept is the same!

## Renaming Columns {.callout-tip}

Use `rename()` within a pipeline:

```{r}
#| eval: false
#| echo: true

tidy_gdp <- year_country_gdp %>% 
  rename(gdp_per_capita = gdpPercap)

head(tidy_gdp)
```

Syntax: `rename(new_name = old_name)`

## Using filter()

Keep only rows that meet criteria:

```{r}
#| eval: false
#| echo: true

year_country_gdp_euro <- gapminder %>%
  filter(continent == "Europe") %>%
  select(year, country, gdpPercap)

head(year_country_gdp_euro)
```

## Combining filter() Conditions

Filter with multiple conditions:

```{r}
#| eval: false
#| echo: true

europe_lifeExp_2007 <- gapminder %>%
  filter(continent == "Europe", year == 2007) %>%
  select(country, lifeExp)

head(europe_lifeExp_2007)
```

## Challenge 1 {.challenge-title}

Write a single command (which can span multiple lines and includes pipes) that will produce a data frame that has the African values for `lifeExp`, `country` and `year`, but not for other Continents.

How many rows does your data frame have and why?

## Challenge 1 Solution {.challenge-title}

```{r}
#| eval: false
#| echo: true

year_country_lifeExp_Africa <- gapminder %>%
  filter(continent == "Africa") %>%
  select(year, country, lifeExp)

nrow(year_country_lifeExp_Africa)
```

624 rows (52 African countries × 12 years of data)

## Order of Operations {.callout-warning}

The order of operations matters!

If we used `select()` first, `filter()` would not be able to find the variable `continent` since we would have removed it.

Always filter before selecting when you need to filter on a column you won't keep!

## Using group_by()

Instead of filtering one group at a time, use `group_by()` to work with all groups:

```{r}
#| eval: false
#| echo: true

str(gapminder)
```

## group_by() Structure

```{r}
#| eval: false
#| echo: true

str(gapminder %>% group_by(continent))
```

Notice the `grouped_df` class - it's now organized by continent!

## Using summarize()

Combine `group_by()` with `summarize()` to calculate statistics per group:

```{r}
#| eval: false
#| echo: true

gdp_bycontinents <- gapminder %>%
  group_by(continent) %>%
  summarize(mean_gdpPercap = mean(gdpPercap))

gdp_bycontinents
```

## Challenge 2 {.challenge-title}

Calculate the average life expectancy per country.

Which has the longest average life expectancy and which has the shortest average life expectancy?

## Challenge 2 Solution {.challenge-title}

```{r}
#| eval: false
#| echo: true

lifeExp_bycountry <- gapminder %>%
  group_by(country) %>%
  summarize(mean_lifeExp = mean(lifeExp))

lifeExp_bycountry %>%
  filter(mean_lifeExp == min(mean_lifeExp) | mean_lifeExp == max(mean_lifeExp))
```

## Challenge 2 Solution (using arrange) {.challenge-title}

Another way using `arrange()`:

```{r}
#| eval: false
#| echo: true

# Shortest life expectancy
lifeExp_bycountry %>%
  arrange(mean_lifeExp) %>%
  head(1)

# Longest life expectancy
lifeExp_bycountry %>%
  arrange(desc(mean_lifeExp)) %>%
  head(1)
```

## Grouping by Multiple Variables

Group by more than one variable:

```{r}
#| eval: false
#| echo: true
#| message: false

gdp_bycontinents_byyear <- gapminder %>%
  group_by(continent, year) %>%
  summarize(mean_gdpPercap = mean(gdpPercap))

head(gdp_bycontinents_byyear)
```

## Multiple Summary Statistics

Create multiple summary variables at once:

```{r}
#| eval: false
#| echo: true
#| message: false

gdp_pop_bycontinents_byyear <- gapminder %>%
  group_by(continent, year) %>%
  summarize(
    mean_gdpPercap = mean(gdpPercap),
    sd_gdpPercap = sd(gdpPercap),
    mean_pop = mean(pop),
    sd_pop = sd(pop)
  )

head(gdp_pop_bycontinents_byyear)
```

## count() and n()

Count observations per group with `count()`:

```{r}
#| eval: false
#| echo: true

gapminder %>%
  filter(year == 2002) %>%
  count(continent, sort = TRUE)
```

## Using n() in Calculations

Use `n()` for the number of observations in calculations:

```{r}
#| eval: false
#| echo: true

gapminder %>%
  group_by(continent) %>%
  summarize(se_le = sd(lifeExp) / sqrt(n()))
```

## Multiple Summary Statistics Example

Calculate min, max, mean, and standard error:

```{r}
#| eval: false
#| echo: true

gapminder %>%
  group_by(continent) %>%
  summarize(
    mean_le = mean(lifeExp),
    min_le = min(lifeExp),
    max_le = max(lifeExp),
    se_le = sd(lifeExp) / sqrt(n())
  )
```

## Using mutate()

Create new columns with `mutate()`:

```{r}
#| eval: false
#| echo: true
#| message: false

gdp_pop_bycontinents_byyear <- gapminder %>%
  mutate(gdp_billion = gdpPercap * pop / 10^9) %>%
  group_by(continent, year) %>%
  summarize(
    mean_gdpPercap = mean(gdpPercap),
    mean_gdp_billion = mean(gdp_billion)
  )

head(gdp_pop_bycontinents_byyear)
```

## mutate() with ifelse()

Combine `mutate()` with `ifelse()` for conditional values:

```{r}
#| eval: false
#| echo: true
#| message: false

gdp_pop_above25 <- gapminder %>%
  mutate(gdp_billion = ifelse(lifeExp > 25, gdpPercap * pop / 10^9, NA)) %>%
  group_by(continent, year) %>%
  summarize(mean_gdp_billion = mean(gdp_billion))

head(gdp_pop_above25)
```

## Scaling Values Conditionally

Update values based on conditions:

```{r}
#| eval: false
#| echo: true
#| message: false

gdp_future <- gapminder %>%
  mutate(gdp_futureExpectation = ifelse(lifeExp > 40, gdpPercap * 1.5, gdpPercap)) %>%
  group_by(continent, year) %>%
  summarize(
    mean_gdpPercap = mean(gdpPercap),
    mean_gdpPercap_expected = mean(gdp_futureExpectation)
  )

head(gdp_future)
```

## Combining dplyr and ggplot2

Load ggplot2:

```{r}
#| eval: false
#| echo: true
#| message: false

library(ggplot2)
```

## Traditional Approach

Creating an intermediate variable:

```{r}
#| eval: false
#| echo: true

americas <- gapminder[gapminder$continent == "Americas", ]

ggplot(data = americas, mapping = aes(x = year, y = lifeExp)) +
  geom_line() +
  facet_wrap(~ country) +
  theme(axis.text.x = element_text(angle = 45))
```

## Piping into ggplot2

No intermediate variable needed:

```{r}
#| eval: false
#| echo: true

gapminder %>%
  filter(continent == "Americas") %>%
  ggplot(mapping = aes(x = year, y = lifeExp)) +
  geom_line() +
  facet_wrap(~ country) +
  theme(axis.text.x = element_text(angle = 45))
```

## Combining mutate and ggplot2

Create new columns and plot in one pipeline:

```{r}
#| eval: false
#| echo: true

gapminder %>%
  mutate(startsWith = substr(country, 1, 1)) %>%
  filter(startsWith %in% c("A", "Z")) %>%
  ggplot(aes(x = year, y = lifeExp, colour = continent)) +
  geom_line() +
  facet_wrap(vars(country)) +
  theme_minimal()
```

## Advanced Challenge {.challenge-title}

Calculate the average life expectancy in 2002 of 2 randomly selected countries for each continent.

Then arrange the continent names in reverse order.

**Hint:** Use the `dplyr` functions `arrange()` and `sample_n()`.

## Advanced Challenge Solution {.challenge-title}

```{r}
#| eval: false
#| echo: true
#| message: false

set.seed(42)  # For reproducibility

lifeExp_2countries_bycontinents <- gapminder %>%
  filter(year == 2002) %>%
  group_by(continent) %>%
  sample_n(2) %>%
  summarize(mean_lifeExp = mean(lifeExp)) %>%
  arrange(desc(mean_lifeExp))

lifeExp_2countries_bycontinents
```

## Key dplyr Verbs Summary

| Verb | Purpose |
|------|---------|
| `select()` | Choose columns |
| `filter()` | Choose rows |
| `group_by()` | Group data |
| `summarize()` | Calculate summaries |
| `mutate()` | Create new columns |
| `arrange()` | Sort rows |
| `count()` | Count observations |
| `rename()` | Rename columns |

## Key Points

-   Use the `dplyr` package to manipulate data frames
-   Use `select()` to choose variables from a data frame
-   Use `filter()` to choose data based on values
-   Use `group_by()` and `summarize()` to work with subsets of data
-   Use `mutate()` to create new variables
-   Use pipes (`%>%`) to chain operations together

## Important Reminders

-   Order of operations matters when combining `select()` and `filter()`
-   `group_by()` changes the data frame structure
-   `n()` returns the count of observations in the current group
-   Pipes make code more readable and reduce intermediate variables
-   Combine dplyr with ggplot2 for powerful data visualization workflows

## Resources

-   [R for Data Science](https://r4ds.hadley.nz/) (online book)
-   [Data Wrangling Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) (PDF)
-   [Introduction to dplyr](https://dplyr.tidyverse.org/) (documentation)
-   [Tidyverse Skills for Data Science](https://jhudatascience.org/tidyversecourse/) (online book)

## Questions?
