---
title: "Subsetting Data in R"
author: "Samuel Bharti"
---

## Overview

Today we will learn to:

-   Subset vectors using indices, names, and logical operations
-   Skip and remove elements from data structures
-   Subset matrices, lists, and data frames
-   Handle special values (NA, NaN, Inf)
-   Combine logical conditions for complex subsetting

## Questions

-   How can I work with subsets of data in R?
-   What are the different ways to extract data?
-   How do I use logical operations for subsetting?

## Creating a Sample Vector

Let's start with a simple numeric vector:

```{r}
#| eval: true
#| echo: true

x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
x
```

## Atomic Vectors {.callout-note}

In R, simple vectors containing:

-   Character strings
-   Numbers
-   Logical values

are called **atomic vectors** because they can't be further simplified.

## Accessing Elements by Index

Extract elements using their position (1-indexed):

```{r}
#| eval: true
#| echo: true

x[1]  # First element
```

```{r}
#| eval: true
#| echo: true

x[4]  # Fourth element
```

The `[]` operator is a function that extracts elements.

## Multiple Elements at Once

Extract several elements by combining indices:

```{r}
#| eval: true
#| echo: true

x[c(1, 3)]  # First and third elements
```

## Slicing Vectors

Use the `:` operator to create sequences:

```{r}
#| eval: true
#| echo: true

x[1:4]  # Elements 1 through 4
```

The `:` operator creates: `c(1, 2, 3, 4)`

## Repeating Elements

You can ask for the same element multiple times:

```{r}
#| eval: true
#| echo: true

x[c(1, 1, 3)]
```

## Out of Bounds

Asking for an index beyond the vector length returns NA:

```{r}
#| eval: true
#| echo: true

x[6]  # Vector only has 5 elements
```

## The Zero Index

Asking for the 0th element returns an empty vector:

```{r}
#| eval: true
#| echo: true

x[0]
```

## Vector Numbering in R {.callout-warning}

**Important:** In R, vector indexing starts at 1, not 0!

-   In C and Python: first element is index 0
-   In R: first element is index 1

This is a common source of confusion for programmers from other languages.

## Skipping Elements

Use negative indices to exclude elements:

```{r}
#| eval: true
#| echo: true

x[-2]  # Everything except element 2
```

## Skipping Multiple Elements

Exclude several elements at once:

```{r}
#| eval: true
#| echo: true

x[c(-1, -5)]  # or x[-c(1, 5)]
```

## Order of Operations {.callout-warning}

**Common mistake** - negating a sequence:

```{r}
#| eval: false
#| echo: true

x[-1:3]  # Error!
```

```         
Error: only 0's may be mixed with negative subscripts
```

The `:` operator runs first: `-1:3` creates `c(-1, 0, 1, 2, 3)`

## Correct Way to Skip Slices

Wrap the sequence in parentheses:

```{r}
#| eval: true
#| echo: true

x[-(1:3)]  # Skip elements 1 through 3
```

Now the `-` operator applies to the entire sequence.

## Removing Elements Permanently

Assign the result back to the variable:

```{r}
#| eval: true
#| echo: true

x <- x[-4]  # Remove element 4
x
```

## Challenge 1 {.challenge-title}

Given the following code:

```{r}
#| eval: false
#| echo: true

x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
```

Come up with at least **2 different commands** that will produce:

```         
  b   c   d
6.2 7.1 4.8
```

Compare your solutions with your neighbor!

## Challenge 1 Solution {.challenge-title}

Multiple approaches work:

```{r}
#| eval: false
#| echo: true

x[2:4]              # By index range
x[-c(1, 5)]         # By skipping indices
x[c("b", "c", "d")] # By name
```

## Subsetting by Name

Extract elements using their names:

```{r}
#| eval: true
#| echo: true

x <- c(a=5.4, b=6.2, c=7.1, d=4.8, e=7.5)
x[c("a", "c")]
```

This is more **reliable** than using positions!

## Why Use Names? {.callout-tip}

Names are safer than indices because:

-   Element positions can change during operations
-   Names remain constant
-   Code is more readable and self-documenting

Always prefer subsetting by name when possible.

## Subsetting with Logical Vectors

Use TRUE/FALSE vectors to select elements:

```{r}
#| eval: true
#| echo: true

x[c(FALSE, FALSE, TRUE, FALSE, TRUE)]
```

Only elements with TRUE are selected.

## Logical Operations for Subsetting

Comparison operators create logical vectors:

```{r}
#| eval: true
#| echo: true

x > 7  # Creates logical vector
```

```{r}
#| eval: true
#| echo: true

x[x > 7]  # Use it to subset
```

## Breaking Down Logical Subsetting

Step by step:

```{r}
#| eval: true
#| echo: true

x > 7  # 1. Evaluate condition
```

This creates: `c(FALSE, FALSE, TRUE, FALSE, TRUE)`

Then R selects elements where TRUE.

## Using == for Subsetting

Find exact matches:

```{r}
#| eval: true
#| echo: true

names(x) == "a"
```

```{r}
#| eval: true
#| echo: true

x[names(x) == "a"]
```

**Remember:** Use `==` for comparison, not `=`

## Combining Logical Conditions

Use `&` (AND) and `|` (OR) to combine conditions:

```{r}
#| eval: true
#| echo: true

x > 5 & x < 7  # Both conditions must be TRUE
```

```{r}
#| eval: true
#| echo: true

x[x > 5 & x < 7]
```

## Logical Operators {.callout-note}

-   `&` - Logical AND: both must be TRUE
-   `|` - Logical OR: either can be TRUE
-   `!` - Logical NOT: inverts TRUE/FALSE

**Note:** Avoid `&&` and `||` in data analysis (they only check first element)

## The NOT Operator

Negate conditions with `!`:

```{r}
#| eval: true
#| echo: true

!(x > 7)  # Invert the logical vector
```

```{r}
#| eval: true
#| echo: true

x[!(x > 7)]  # Elements NOT greater than 7
```

## all() and any()

Check entire vectors:

```{r}
#| eval: true
#| echo: true

all(x > 4)  # Are ALL elements > 4?
```

```{r}
#| eval: true
#| echo: true

any(x > 7)  # Are ANY elements > 7?
```

## Challenge 2 {.challenge-title}

Given:

```{r}
#| eval: false
#| echo: true

x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
```

Write a subsetting command to return values in x that are **greater than 4 AND less than 7**.

## Challenge 2 Solution {.challenge-title}

```{r}
#| eval: false
#| echo: true

x[x > 4 & x < 7]
```

Both conditions must be TRUE for an element to be included.

## Non-Unique Names {.callout-warning}

Multiple elements can have the same name:

```{r}
#| eval: true
#| echo: true

x <- 1:3
names(x) <- c('a', 'a', 'a')
x['a']  # Only returns first value
```

```{r}
#| eval: true
#| echo: true

x[names(x) == 'a']  # Returns all three
```

## Getting Help for Operators {.callout-tip}

Wrap operators in quotes to search for help:

```{r}
#| eval: false
#| echo: true

help("%in%")
?"%in%"
```

## Skipping Named Elements

Can't use negative with names:

```{r}
#| eval: false
#| echo: true

x <- c(a=5.4, b=6.2, c=7.1, d=4.8, e=7.5)
x[-"a"]  # Error!
```

```         
Error: invalid argument to unary operator
```

## Using != to Skip Names

Use the "not equals" operator instead:

```{r}
#| eval: true
#| echo: true

x <- c(a=5.4, b=6.2, c=7.1, d=4.8, e=7.5)
x[names(x) != "a"]
```

## Skipping Multiple Names - Wrong Way

This seems logical but has a problem:

```{r}
#| eval: true
#| echo: true

x[names(x) != c("a", "c")]
```

**Warning:** Longer object length is not a multiple of shorter object length

Element "c" is still included - not what we wanted!

## Understanding Recycling

What does `!=` actually do?

```{r}
#| eval: true
#| echo: true

names(x) != c("a", "c")
```

R **recycles** the shorter vector: `c("a", "c", "a", "c", "a")`

This creates incorrect comparisons!

## Recycling Visualization {.callout-warning}

```
names(x):   a    b    c    d    e
compared:   a    c    a    c    a  (recycled)
result:   FALSE TRUE TRUE TRUE TRUE
```

Element 3 ("c") compared to "a" → TRUE (wrong!)

## The Correct Way: %in%

Use the `%in%` operator for multiple matches:

```{r}
#| eval: true
#| echo: true

names(x) %in% c("a", "c")
```

```{r}
#| eval: true
#| echo: true

x[!names(x) %in% c("a", "c")]  # Use ! to exclude
```

## How %in% Works {.callout-note}

The `%in%` operator asks: "Does this element occur in the second vector?"

-   Goes through each element of left vector
-   Checks if it exists anywhere in right vector
-   Returns TRUE/FALSE for each element
-   No recycling problems!

## Challenge 3 {.challenge-title}

Southeast Asia countries:

```{r}
#| eval: false
#| echo: true

seAsia <- c("Myanmar","Thailand","Cambodia","Vietnam","Laos")
gapminder <- read.csv("data/gapminder_data.csv")
countries <- unique(as.character(gapminder$country))
```

Create a logical vector that is TRUE for southeast Asian countries.

Come up with 3 approaches:

1.  Wrong way (using only `==`)
2.  Clunky way (using `==` and `|`)
3.  Elegant way (using `%in%`)

## Challenge 3 Solution {.challenge-title}

```{r}
#| eval: false
#| echo: true

# Wrong - gives warning
countries == seAsia

# Clunky - works but tedious
countries == "Myanmar" | countries == "Thailand" | 
  countries == "Cambodia" | countries == "Vietnam" | 
  countries == "Laos"

# Elegant - best approach
countries %in% seAsia
```

## Handling Special Values

R has special functions for dealing with missing/invalid data:

-   `is.na()` - finds NA or NaN
-   `is.nan()` - finds NaN only
-   `is.infinite()` - finds Inf
-   `is.finite()` - finds normal values (excludes NA, NaN, Inf)
-   `na.omit()` - removes all missing values

## Factor Subsetting

Factors work like vectors:

```{r}
#| eval: true
#| echo: true

f <- factor(c("a", "a", "b", "c", "c", "d"))
f[f == "a"]
```

```{r}
#| eval: true
#| echo: true

f[f %in% c("b", "c")]
```

## Factors Keep All Levels

Skipping elements doesn't remove levels:

```{r}
#| eval: true
#| echo: true

f[-3]  # Removed "b" value
```

Notice: Levels still shows all 4 levels (a, b, c, d)

## Matrix Subsetting

Matrices use `[row, column]` notation:

```{r}
#| eval: true
#| echo: true

set.seed(1)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m[3:4, c(3,1)]  # Rows 3-4, columns 3 and 1
```

## Selecting All Rows or Columns

Leave an argument blank to get all:

```{r}
#| eval: true
#| echo: true

m[, c(3,4)]  # All rows, columns 3-4
```

## Matrix to Vector Conversion

Single row/column becomes a vector:

```{r}
#| eval: true
#| echo: true

m[3, ]  # Returns a vector
```

## Preserving Matrix Structure

Use `drop = FALSE` to keep as matrix:

```{r}
#| eval: true
#| echo: true

m[3, , drop=FALSE]  # Still a matrix
```

## Matrix Error Handling

Out-of-bounds access throws an error:

```{r}
#| eval: false
#| echo: true

m[, c(3, 6)]  # Only 4 columns!
```

```         
Error: subscript out of bounds
```

Matrices are stricter than vectors.

## Higher Dimensional Arrays {.callout-note}

For 3D arrays:

-   First argument = rows
-   Second argument = columns
-   Third argument = depth

Each dimension gets its own argument in `[]`.

## Matrices as Vectors

Matrices can be accessed with single index:

```{r}
#| eval: true
#| echo: true

m[5]  # 5th element in column-major order
```

## Column-Major Format {.callout-note}

Matrices are stored column-wise by default:

```{r}
#| eval: true
#| echo: true

matrix(1:6, nrow=2, ncol=3)
```

Elements fill down columns first, then across.

## Row-Major Format

Use `byrow=TRUE` to fill by rows:

```{r}
#| eval: true
#| echo: true

matrix(1:6, nrow=2, ncol=3, byrow=TRUE)
```

## Challenge 4 {.challenge-title}

Given:

```{r}
#| eval: false
#| echo: true

m <- matrix(1:18, nrow=3, ncol=6)
print(m)
```

Which command extracts values 11 and 14?

A. `m[2,4,2,5]`  
B. `m[2:5]`  
C. `m[4:5,2]`  
D. `m[2,c(4,5)]`

## Challenge 4 Solution {.challenge-title}

Answer: **D**

```{r}
#| eval: false
#| echo: true

m[2, c(4,5)]  # Row 2, columns 4 and 5
```

-   Row 2, column 4 = 11
-   Row 2, column 5 = 14

## List Subsetting with []

`[` returns a list:

```{r}
#| eval: true
#| echo: true

xlist <- list(a = "Software Carpentry", 
              b = 1:10, 
              data = head(mtcars))
xlist[1]  # Returns list with one element
```

## List Subsetting Multiple Elements

```{r}
#| eval: true
#| echo: true

xlist[1:2]  # Returns list with two elements
```

## Extracting Elements with [[]]

`[[]]` extracts the actual element:

```{r}
#| eval: true
#| echo: true

xlist[[1]]  # Returns the vector itself
```

Now the result is a character vector, not a list!

## [[]] Limitations

Can't extract multiple elements:

```{r}
#| eval: false
#| echo: true

xlist[[1:2]]  # Error!
```

```         
Error: subscript out of bounds
```

Can't skip elements:

```{r}
#| eval: false
#| echo: true

xlist[[-1]]  # Error!
```

## Extracting by Name

Use names with `[[]]`:

```{r}
#| eval: true
#| echo: true

xlist[["a"]]
```

## The $ Shortcut

`$` is shorthand for extracting by name:

```{r}
#| eval: false
#| echo: true

xlist$data
```

Equivalent to `xlist[["data"]]`

## Challenge 5 {.challenge-title}

Given:

```{r}
#| eval: false
#| echo: true

xlist <- list(a = "Software Carpentry", 
              b = 1:10, 
              data = head(mtcars))
```

Extract the number **2** from xlist.

Hint: The number 2 is in the "b" item.

## Challenge 5 Solution {.challenge-title}

```{r}
#| eval: false
#| echo: true

xlist[[2]][2]  # or xlist[["b"]][2] or xlist$b[2]
```

First `[[2]]` extracts the vector, then `[2]` gets the second element.

## Challenge 6 {.challenge-title}

Given a linear model:

```{r}
#| eval: false
#| echo: true

mod <- aov(pop ~ lifeExp, data=gapminder)
```

Extract the **residual degrees of freedom**.

Hint: `attributes()` will help you!

## Challenge 6 Solution {.challenge-title}

```{r}
#| eval: false
#| echo: true

attributes(mod)  # See all available attributes
mod$df.residual
```

## Data Frame Subsetting: Single Argument

`[` with one argument acts on columns:

```{r}
#| eval: false
#| echo: true

gapminder <- read.csv(
  "https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/main/episodes/data/gapminder_data.csv"
)

head(gapminder[3])  # Returns data frame with column 3
```

## Data Frame Subsetting: [[]]

`[[]]` extracts a column as a vector:

```{r}
#| eval: false
#| echo: true

head(gapminder[["lifeExp"]])
```

## Data Frame Subsetting: $

`$` is the convenient shorthand:

```{r}
#| eval: false
#| echo: true

head(gapminder$year)
```

## Data Frame: Two Arguments

With `[row, column]`, acts like a matrix:

```{r}
#| eval: false
#| echo: true

gapminder[1:3, ]  # First 3 rows, all columns
```

## Single Row Subsetting

Single row returns a data frame:

```{r}
#| eval: false
#| echo: true

gapminder[3, ]
```

Mixed types preserved in data frame structure.

## Challenge 7 {.challenge-title}

Fix these common data frame subsetting errors:

1. Extract observations from 1957:
```{r}
#| eval: false
#| echo: true
gapminder[gapminder$year = 1957,]
```

2. Extract all columns except 1 through 4:
```{r}
#| eval: false
#| echo: true
gapminder[, -1:4]
```

## Challenge 7 Continued {.challenge-title}

3. Extract rows where life expectancy > 80:
```{r}
#| eval: false
#| echo: true
gapminder[gapminder$lifeExp > 80]
```

4. Extract first row, columns 4 and 5:
```{r}
#| eval: false
#| echo: true
gapminder[1, 4, 5]
```

5. Extract rows for years 2002 and 2007:
```{r}
#| eval: false
#| echo: true
gapminder[gapminder$year == 2002 | 2007,]
```

## Challenge 7 Solution {.challenge-title}

```{r}
#| eval: false
#| echo: true

# 1. Use == not =
gapminder[gapminder$year == 1957,]

# 2. Wrap range in parentheses
gapminder[, -(1:4)]

# 3. Need comma for rows
gapminder[gapminder$lifeExp > 80, ]

# 4. Use c() for multiple columns
gapminder[1, c(4, 5)]

# 5. Complete both comparisons
gapminder[gapminder$year == 2002 | gapminder$year == 2007,]
# or better:
gapminder[gapminder$year %in% c(2002, 2007),]
```

## Challenge 8 {.challenge-title}

1.  Why does `gapminder[1:20]` return an error? How does it differ from `gapminder[1:20, ]`?

2.  Create a new data frame called `gapminder_small` that only contains rows 1 through 9 and 19 through 23. You can do this in one or two steps.

## Challenge 8 Solution {.challenge-title}

```{r}
#| eval: false
#| echo: true

# 1. gapminder[1:20] tries to get columns 1-20
#    gapminder[1:20, ] gets rows 1-20 (correct)

# 2. One step:
gapminder_small <- gapminder[c(1:9, 19:23), ]

# Two steps:
gapminder_small <- gapminder[1:9, ]
gapminder_small <- rbind(gapminder_small, gapminder[19:23, ])
```

## Key Points

-   Indexing in R starts at 1, not 0
-   Access individual values by location using `[]`
-   Access slices of data using `[low:high]`
-   Access arbitrary sets using `[c(...)]`
-   Use logical operations to access subsets
-   `%in%` is essential for matching multiple values

## Key Points (Continued)

-   Negative indices skip elements
-   `[[]]` extracts list elements, `[]` subsets lists
-   `$` is shorthand for extracting named elements
-   Data frames can be subset like lists or matrices
-   Use `drop=FALSE` to preserve structure

## Common Pitfalls {.callout-warning}

1.  Using `=` instead of `==` for comparison
2.  Forgetting the comma in `[row, column]`
3.  Recycling issues with `!=` (use `%in%` instead)
4.  Forgetting R uses 1-based indexing
5.  Order of operations with `-` and `:`

## Best Practices {.callout-tip}

-   Subset by name when possible (more reliable)
-   Use `%in%` for multiple value matching
-   Check your subset before assigning back
-   Use `str()` to understand data structure
-   Combine conditions with `&` and `|` for clarity

## Practice Makes Perfect

Master these six ways to subset:

1.  Positive integers (position)
2.  Negative integers (exclusion)
3.  Logical vectors (conditions)
4.  Named indices (by name)
5.  Empty (all elements)
6.  Zero (empty result)

## Resources

-   Software Carpentry: [r-novice-gapminder](https://swcarpentry.github.io/r-novice-gapminder/)
-   `?'['` - Help on subsetting
-   `?'%in%'` - Help on matching operator
-   RStudio Cheatsheets

## Questions?
