---
title: "Exploring Data Frames in R"
subtitle: "Adapted from Software Carpentry by Samuel Bharti"
---

## Overview

Today we will learn to:

-   Add and remove rows or columns
-   Append two data frames
-   Explore data frame properties
-   Read data from CSV files

## What is a Data Frame?

A data frame is a table where:

-   Columns are vectors (same data type within column)
-   Rows are lists (can mix data types across columns)
-   Most common data structure in R

## Reading Data

Load data from CSV files:

```{r}
#| eval: true
#| echo: true

cats <- read.csv("data/feline-data.csv")
cats
```


## Adding Columns

Use `cbind()` to add columns:

```{r}
#| eval: true
#| echo: true

age <- c(2, 3, 5)
cbind(cats, age)
```


## Try: Adding a Different Vector

What happens if we try with a different number of values?

```{r}
#| eval: false
#| echo: true

age <- c(2, 3, 5, 12)
cbind(cats, age)
```

## Error: Too Many Values

```         
Error in `data.frame()`:
! arguments imply differing number of rows: 3, 4
```

The data frame has 3 rows but age has 4 values.

## Adding Columns: Key Rule

Number of rows must match vector length:

```{r}
#| eval: false
#| echo: true

nrow(cats)  # 3
length(age) # must also be 3
```

Mismatched lengths will cause an error.

## Saving the New Column

To keep the new column, assign it back to cats:

```{r}
#| eval: true
#| echo: true

age <- c(2, 3, 5)
cats <- cbind(cats, age)
cats
```

## Adding Rows

Use `rbind()` to add rows (as lists):

```{r}
#| eval: true
#| echo: true

newRow <- list("tortoiseshell", 3.3, TRUE, 9)
cats <- rbind(cats, newRow)
cats
```

## Removing Rows

Use negative indices to drop rows:

```{r}
#| eval: true
#| echo: true

# Remove row 4
cats[-4, ]
```

Remove multiple rows:

```{r}
#| eval: true
#| echo: true

# Remove rows 3 and 4
cats[c(-3, -4), ]
```

## Removing Columns

Use negative column indices:

```{r}
#| eval: true
#| echo: true

# Remove column 4
cats[, -4]
```

Or use column names with `%in%`:

```{r}
#| eval: true
#| echo: true

drop <- names(cats) %in% c("age")
cats[, !drop]
```

## Appending Data Frames

Combine two data frames with `rbind()`:

```{r}
#| eval: true
#| echo: true

cats <- rbind(cats, cats)
cats
```

Result: duplicate rows appended to original data

## Challenge 1 {.challenge-title}

Create a data frame with your information:

```{r}
#| eval: false
#| echo: true

df <- data.frame(
  first_name = c("Your", "Name"),
  last_name = c("Goes", "Here"),
  lucky_number = c(7, 13)
)
```

Then:

1.  Use `rbind()` to add an entry for the person sitting beside you

2.  Use `cbind()` to add a column with each person's answer to the question, "Is it time for coffee break?"

## Commit Your Work

Save progress to version control:

``` bash
git add .
git commit -m "Add data frame manipulation examples"
```

## Reading the Gapminder Dataset

Now let's work with a realistic dataset. Load the gapminder data:

```{r}
#| eval: false
#| echo: true

gapminder <- read.csv("data/gapminder_data.csv")
```

Or read directly from the internet:

```{r}
#| eval: true
#| echo: true

gapminder <- read.csv(
  "https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/main/episodes/data/gapminder_data.csv"
)
```

## Exploring Gapminder: str()

Check the structure of the dataset:

```{r}
#| eval: false
#| echo: true

str(gapminder)
```

Output shows 1704 observations and 6 variables:

-   `country` (character)
-   `year` (integer)
-   `pop` (numeric)
-   `continent` (character)
-   `lifeExp` (numeric)
-   `gdpPercap` (numeric)

## Exploring Gapminder: summary()

Get summary statistics:

```{r}
#| eval: true
#| echo: true

summary(gapminder)
```

Shows min, quartiles, median, mean, and max for each column.

## Data Type by Column

Check individual column types:

```{r}
#| eval: true
#| echo: true

typeof(gapminder$year)      # integer
typeof(gapminder$country)   # character
str(gapminder$country)      # character vector length 1704
```

## Checking Data Frame Properties

Examine dimensions and types:

```{r}
#| eval: true
#| echo: true

typeof(gapminder)        # list
length(gapminder)        # 6 (columns)
nrow(gapminder)          # 1704 rows
ncol(gapminder)          # 6 columns
dim(gapminder)           # 1704 6
colnames(gapminder)      # column names
```

## Viewing the Data

Check the first and last rows:

```{r}
#| eval: true
#| echo: true

head(gapminder)
tail(gapminder)
```

## Challenge 2 {.challenge-title}

Examine the gapminder data:

1.  Check the last few lines of the data
2.  Check some rows in the middle
3.  Try to pull a few random rows

(Hint: Use functions like `tail()`, subsetting, and `sample()`)

## Challenge 2 Solution {.challenge-title}

```{r}
#| eval: false
#| echo: true

tail(gapminder)

# Middle rows (example: rows 800–810)
gapminder[800:810, ]

# Random rows
gapminder[sample(nrow(gapminder), 5), ]
```

## Challenge 3 {.challenge-title}

Create an R script:

1.  Go to File \> New File \> R Script
2.  Write code to load the gapminder dataset
3.  Save in the `scripts/` directory
4.  Add to version control
5.  Run the script using `source()` function

## Challenge 4 {.challenge-title}

Interpret the output of `str(gapminder)`:

Using what you know about lists and vectors, explain:

-   What does each line of `str()` output mean?
-   Why is length 6 but nrow is 1704?
-   How does this relate to the data structure?

Discuss with your neighbor.

## Important Reminders

-   Columns must have consistent types
-   Number of elements must match when adding columns
-   Rows are added as lists
-   Always check structure after loading data
-   Use `str()` first, then explore further

## Commit Your Final Work

Save all changes to version control:

``` bash
git add .
git commit -m "Complete data frame exploration lesson"
```

Always commit your final work.

## Resources

-   Software Carpentry: [r-novice-gapminder](https://swcarpentry.github.io/r-novice-gapminder/)
-   RStudio Cheatsheets
-   R Documentation: `?read.csv`, `?str`, etc.

## Questions?
