---
title: "Seeking Help"
subtitle: "Adapted from Software Carpentry"
---

## Overview

**Questions:**

- How can I get help in R?

**Objectives:**

- Use R help files and resources
- Find packages for specific tasks
- Seek help from peers
- Use generative AI responsibly

## Reading Help Files

Access help for any function:

```{r}
#| eval: false
#| echo: true

?function_name
help(function_name)
```

Help pages include: Description, Usage, Arguments, Details, Value, See Also, and Examples.

For special operators, use quotes or backticks:

```{r}
#| eval: false
#| echo: true

?"<-"
?`<-`
```

:::{.callout-tip}
## Running Examples

Highlight code examples in help pages and hit Ctrl+Return to run them in RStudio.
:::

## Getting Help

### Package Vignettes

Packages often include tutorials and extended examples:

```{r}
#| eval: false
#| echo: true

vignette()                    # list all vignettes
vignette(package="dplyr")     # list dplyr vignettes
vignette("dplyr")             # open specific vignette
```

See [RStudio cheatsheets](https://rstudio.com/resources/cheatsheets/) for common packages.

### Fuzzy Search

Search for functions when you don't know the exact name:

```{r}
#| eval: false
#| echo: true

??function_name
```

### Finding Packages

[CRAN Task Views](https://cran.r-project.org/web/views) organizes packages by topic.

## Getting Help from Others

### Online Resources

- **Search the internet**: Paste your error message and "R" into a search engine
  - [StackOverflow](https://stackoverflow.com/questions/tagged/r) is particularly helpful; use the `[r]` tag
  - Note: Understand code before copying it
  
- **Ask colleagues**: Share your problem with someone more experienced
- **Rubber duck debugging**: Explaining your problem out loud often reveals the solution

### Formulating Good Questions

When asking for help, provide:

```{r}
#| eval: false
#| echo: true

dput(your_data)        # share your data reproducibly
sessionInfo()          # share your R environment
```

## Generative AI Tools

Tools like ChatGPT can provide helpful suggestions. However:

- Responses are **not always reliable** - verify accuracy
- They may generate plausible-sounding but incorrect information
- You need domain knowledge to evaluate code quality
- Later, focus on building your own problem-solving skills

::: {.callout-note icon=false}
## Challenge 1: Understanding the c Function

Look at the help page for the `c` function. What kind of vector do you expect will be created if you evaluate the following:

```r
c(1, 2, 3)
c('d', 'e', 'f')
c(1, 2, 'f')
```

::: {.callout-tip collapse="true"}
## Solution to Challenge 1

The `c()` function creates a vector, and all elements of a vector must be the same type. In the first two cases, the elements are already the same type. But in the third case, R will convert all the numbers to characters to make them all the same type.

```r
c(1, 2, 3)      # numeric vector
c('d', 'e', 'f')  # character vector
c(1, 2, 'f')    # character vector: "1" "2" "f"
```
:::
:::

::: {.callout-note icon=false}
## Challenge 2: Understanding paste

Look at the help for the `paste` function. You will need to use it later. What's the difference between the `sep` and `collapse` arguments?

::: {.callout-tip collapse="true"}
## Solution to Challenge 2

`sep` controls how elements are separated within a single call to paste, while `collapse` controls how multiple results are combined into a single string.

```r
paste("a", "b", sep = "-")         # "a-b"
paste(c("a", "b"), c("c", "d"), sep = "-")  # "a-c" "b-d"
paste(c("a", "b"), c("c", "d"), sep = "-", collapse = " & ")  # "a-c & b-d"
```
:::
:::

::: {.callout-note icon=false}
## Challenge 3: Finding Functions for Reading Data

Use help to find a function (and its associated parameters) that you could use to load data from a tabular file in which columns are delimited with "\t" (tab) and the decimal point is a "." (period). This check for decimal separator is important, especially if you are working with international colleagues, because different countries have different conventions for the decimal point (i.e. comma vs period). 

**Hint**: use `??"read table"` to look up functions related to reading in tabular data.

::: {.callout-tip collapse="true"}
## Solution to Challenge 3

The standard R function for reading tab-delimited files with a period as decimal separator is `read.delim()`. You can also use `read.table()` with appropriate arguments:

```r
read.delim(file, header = TRUE, sep = "\t", dec = ".")
# or
read.table(file, header = TRUE, sep = "\t", dec = ".")
```
:::
:::

## Resources

- [Quick R](https://www.statmethods.net/)
- [RStudio cheat sheets](https://www.rstudio.com/resources/cheatsheets/)
- [Cookbook for R](https://www.cookbook-r.com/)

## Key Points

- Use `help()` or `?` to get online help in R
- Use `??` for fuzzy searches when you don't know the exact function name
- R help files provide detailed documentation including usage, arguments, and examples
- CRAN Task Views can help you find packages for specific tasks
- When asking for help, provide reproducible examples and session information
